Analysis of the 2005--\myrevision{2016} Earthquake Sequence in Northern Iran Using the Visibility Graph Method

### 2017-06-22 response.tex

% ******************************************************************
% *************************** REVIEWER 1 ***************************
% ******************************************************************

\begin{center}
	\bf
	\large
	Authors' Response to Reviewer 1
\end{center}

\noindent
We thank the reviewer for his/her comments and suggestions, which helped us improve the manuscript. To help the review process, in the annotated manuscript version of the paper, we have highlighted the most relevant changes that resulted from these comments in green font. In the following, we provide a response to each comment. The original comments are in italic black font, followed by our responses in regular blue font.
\vspace{2ex}
\newline

\introcomment{~}{%
This is an interesting paper that applies a relatively new technique to the seismicity data in Northern Iran. The manuscript is well-written and would be a worthy contribution to the Pageoph journal. I have a few moderate-to-major concerns that I recommend to be addressed by the authors in a revised manuscript.
}

\comment{1}{%
Please explain why you decluster the catalog and be more specific on the declustering method/declustering-parameters and the results of the declustering procedure. How many events are in the catalog before/after declustering?
}

\response{%
Application of VG approach in the seismic catalog is a novel approach. Recent studies have used both declustered (e.g., Telesca et al (2014b) and Telesca et al (2016)) and the whole catalog (Telesca and Lovallo (2012) and Telesca et al (2013)). Based on this comment and subsequent comment of the reviewer (and also the comments of second reviewers), we consider both whole and declustered catalog. We used Gardner and Knopoff (1974) method for declustering the catalog and reported the catalog size for original data (whole catalog), whole catalog after considering completeness magnitude and also declustered catalog in Table 1.  
}

\comment{2}{%
In order to check how well the declustering method worked for the analyzed datasets, would be useful to show the earthquakes in a plot of longitude versus time.
}

\response{%
That is a good point. However, we think it gives an idea of distance based analysis of VG which is far from the scope of this article. We add the figure of original catalog before and after considering the completeness magnitude (whole and complete catalogs.)
}

\comment{3}{%
I think would be interesting to check/discuss/compare the k-M slope and b-value relation for both the original and the declustered catalogs.
}

\response{%
Thanks for the comment. We considered both catalogs and presented the results. Please see Fig. 7. 
}

\comment{4}{%
I would suggest to check the results for several magnitude thresholds above the completeness magnitude. Are the conclusions robust or do they depend on the chosen magnitude thresholds?
}

\response{%
The seismicity parameters change based on completeness magnitude. The parameters which come from the VG analysis also follow this pattern. According to this comment and comment \#3 of second reviewer we compute time dependent completeness test and choose a conservative completeness magnitude.  
}

\comment{5}{%
Would it be possible to calculate a correlation coefficient for the $k-M$ slopes and the $b-values$ in Figure 9 (middle plot)?
}

\response{%
We can add the numbers into figure 10. \\
 (Dec catalog:\ \ \ \   $ k_m :  Az =  4.4\%  Al =  7.1 \%  Ko =  6.6\% -- b: Az = 6.1\%  Al = 3.3\% Ko=4.8\%$)\\
 (Whole catalog: $ k_m :  Az =  9.4\%  Al =  7.4 \%  Ko =  6.2\% -- b: Az = 5.4\%  Al = 4.8\% Ko=5.5\%$)
}

\comment{6}{%
The authors use 20 events (in a moving window) for estimating the b-values and k-M slopes in Figure 9. I think this is an extremely small number and I am concerned about the reliability of the b-value estimates. At least for the areas of higher seismicity the authors should also present graphs obtained using a larger number of events per moving window.
}

\response{%
Thanks for the comment. We tried different window size from 20 to 100. As it is natural to expect, there are trade-offs between different window sizes. Smaller window sizes emphasize the local changes in the time series, but are less reliable when it comes to the seismicity parameters (e.g., b value). Larger window sizes, on the other hand, offer less insight into the time dependence of the window sequence characteristics. We chose a window size $n=50$, which is consistent with the suggested minimum number of events acceptable to estimate b according to  Woessner and Wiemer (2005). We modified the text accordingly. 
}

\comment{7}{%
The discussion of b-values in relation with larger earthquakes (page 10) in Figure 9 is unclear. From a quick visual inspection, I fail to see any distinct anomalies before larger (M > 5.0) events. I think that a lack of correlation should be also discussed in the manuscript, if true.
}

\response{%
The correlation for 2012 Mw 6.4 Azerbaijan earthquake is clear to see. However, we agree with the reviewer that the drops of b-value is not consistent. Or we should say in the case of big earthquake there is always a drop in b and k-m value, however, there are some drops in the window analysis which is not aligned with big earthquake. We addressed the lack of correlation for the KopehDagh region. We also discussed that the lack of correlation could be the results of relatively high completeness magnitude. The text is modified accordingly.
}

@@

\newpage

\begin{center}
	\bf
	\large
	Authors' Response to Reviewer 2
	\end{center}

\noindent
We thank the reviewer for his/her comments and suggestions, which helped us improve the manuscript. To help the review process, in the annotated manuscript version of the paper, we have highlighted the most relevant changes that resulted from these comments in green font. In the following, we provide a response to each comment. The original comments are in italic black font, followed by our responses in regular blue font.
\vspace{2ex}
\newline

\introcomment{~}{%
The authors present an analytical method to estimate statistical properties of earthquake catalogs and they apply their method in Iranian catalog. Although I find some promise in the method I suggest that more analysis is required for the study to be comprehensive. 
}

\comment{1}{%
1) Declustering the earthquake catalog, especially using parameters derived for California seismicity under modest monitoring conditions only brings additional uncertainty in the analysis. I am not sure why the authors believe that they should declutter the catalog. They could report if they find any time variability of b-value before/after aftershock sequences. i would suggest not going ahead with declustering.
}

\response{%
Thanks for the comment. We consider both catalogs.
}

\comment{2}{%
 Uncertainty analysis for standard b-value estimation for sample population, magnitude correction
}

\response{
Thanks. We conduct the uncertainty analysis according to Shi and Bolt 1982. Please see equation 3 and Table 1. 
}

\comment{3}{%
I find that Mc (magnitude of completeness) determination, should be done in a time-dependent manner since from the Fig. 5 it is obvious that monitoring conditions change.
}

\response{
That is right. We conduct a time dependent Mc estimation through Maximum Curvature method implemented in ZMAP software. Please see Fig. 4.  The results modified accordingly. 
}

\comment{4}{%
How the Authors account for Bam earthquake in 2003 that has aftershock in the early part of their catalog?
}

\response{
Our catalog starts from 2005. We agree the fact that  the earthquake could affect the seismicity pattern several consequent years, however, the spatial distance of the epicenter from the study region is fairly large ($d > 400 miles$). Therefore, we believe $2003 Mw 6.6$ Bam earthquake's aftershocks is not included in the current study regions' catalog.  
}

### 2017-06-22 methodology.tex

Discrete mathematics and computer science often deal with different types of data abstractions and structures such as sets, trees, and graphs, where concepts like nodes and links are used to describe the topology of a collection of objects and manipulate their data \citep[e.g.,][]{Skiena_2008_Book}. In graphs, the data---sometimes also referred to as the payload---resides at the vertices or nodes who are connected through links or edges, which may or may not have an associated direction. A visibility graph is a special type of undirected graphs in which the links are straight lines connecting intervisible nodes; that is, straight lines that do not go across any obstacle while connecting nodes that can see each other in a physical space \citep{LozanoPerez_1979_CACM}. Visibility graphs have been mostly used in robotics for navigation path planning \citep[e.g.,][]{Huang_2004_Proc, Oommen_1987_JRA} but have also seen applications in other fields including urban studies, interior architecture, medicine, and geosciences \citep[e.g.,][]{Raman_2010_UE, Ahmadlou_2010_JNT, Varoudis_2014_JSS, Phillips_2015_ESR}.

>>>>

Discrete mathematics and computer science often deal with different types of data abstractions and structures such as sets, trees, and graphs, where concepts like nodes and links are used to describe the topology of a collection of objects and manipulate their data \citep[e.g.,][]{Skiena_2008_Book}. In graphs, the data---sometimes also referred to as the payload---resides at the vertices or nodes who are connected through links or edges, which may or may not have an associated direction. A visibility graph is a special type of graphs in which the links are straight lines connecting intervisible nodes; that is, straight lines that do not go across any obstacle while connecting nodes that can see each other in a physical space \citep{LozanoPerez_1979_CACM}. Visibility graphs have been mostly used in robotics for navigation path planning \citep[e.g.,][]{Huang_2004_Proc, Oommen_1987_JRA} but have also seen applications in other fields including urban studies, interior architecture, medicine, and geosciences \citep[e.g.,][]{Raman_2010_UE, Ahmadlou_2010_JNT, Varoudis_2014_JSS, Phillips_2015_ESR}.

### 2017-06-22 results.tex

We generate the visibility graphs for each subregion of interest using the time-magnitude sequences \myrevision{of both the complete and declustered catalogs. In the case of the complete catalog, that corresponds to the sequences shown in Fig.~\ref{fig:mag-time}. In every case, the numbers of nodes in each graph are the same as those shown in Table \ref{tab:seismicity}; and} the links between inter-visible events are established following the condition in equation (\ref{eq:vg}), as in the example shown in Fig.~\ref{fig:vg}. (We do not include a visualization of the graphs because the links are so many, that it is only practical to visualize the graphs of short sequences.) We collect information about the number of inter-visibility links associated with each event (connectivity degree, $k$) and categorize the events in magnitude bins of size $\Delta M_{\mathrm{bin}} = 0.1$, as mentioned in the Methodology section. Figure \ref{fig:km} shows the scattered distribution of events in the magnitude-connectivity degree plane for each subregion\myrevision{, corresponding to the complete catalogs.} The figure shows that, in general, the value of $k$ increases with $M_w$\myrevision{; and includes the results of obtaining linear $k$-$M$ regressions for each dataset along with the values of the slope $m$ for the three seismic zones, namely 10.0, 9.85, and 8.70 for the Azerbaijan, Alborz, and Kopeh Dagh regions. Similar results regressions were obtained for the declustered catalogs, which yielded $m$ values of 8.92, 8.97, and 7.96, respectively.}

@@

Next, we examine the relationship between the $b$ values from Table \ref{tab:seismicity} and the $k$-$M$ slopes. \myrevision{Figures \ref{fig:regression}a and \ref{fig:regression}a show the scattered results of the $m$ versus $b$ for the complete and declustered catalogs, respectively. Along with the data-points obtained for our analysis of the three seismic regions of northern Iran, in each plot, we also include the data points corresponding to the magnitude-time sequences of the Mexican subduction zone provided by \citet{Telesca2013}, the Pannoninan seismic zone from \citet{Telesca2014}, the experimental results obtained by \citet{Telesca2014-pone}, and the result from a similar preliminary analysis done by \citet{Azizzadeh_2017_SSA} for California. These figures also include the linear trends of the data points, as they are aggregated (i.e., each regression reflects the addition of new data-points from different studies). In the figure, we indicate which data points were considered for each of the computed linear regressions and their corresponding correlation coefficient, $R$.}

@@

To further explore the sensitivity of the graph properties to the number of events in each catalog and the value of the minimum magnitude, we randomly picked a significant number of subsequences from within the initial catalog compiled for each region, and repeated the analysis for each subsequence. In total, for each region, we extracted 200 new subsequences from the \myrevision{whole} catalog. The number of events in each subsequence, $n$, was varied randomly but chosen to be large enough to represent the seismic characteristics of each region. In particular, the minimum size of each subsequence was set to be $n \geq 200$, and the maximum size in the sequence was set to be as large as the original \myrevision{whole} catalog (see Table \ref{tab:seismicity}). We forced the random subsequences to progress positively in time without altering the natural occurrence of events. In other words, we randomly determined the initial event and the subsequence size (number of events to be considered), and then picked that number of events following the initial earthquake in the subsequence. Next, we determined the value of $M_c$ and $b$ \myrevision{to obtain the complete and declustered catalogs for each region}, created the graph for all events with $M \geq M_c$, and extracted the connectivity degree of the events in each subsequence. \myrevision{Because this needed to be done for a large number of subsequences, in this part of the analysis we used the goodness-of-fit test (GFT) method to determine $M_c$ systematically, instead of the MAXC method. The GFT method follows the recommendations of \citet{Wiemer2000} and \citet{Wiemer2001}, and leads to less conservative values of $M_c$, which is preferable to keep the subsequences large enough.}

\begin{figure*}%[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-08} 
	\caption{\myrevision{Correlation between the $k$-$M$ slope ($m$) and $b$ values for 200 random sequences extracted from the complete (a) and declustered (d) catalogs of the three northern Iranian seismic regions considered in this study (scattered symbols), including the mean values (empty symbols with thick border) and the data points obtained from the regional analyses (solid symbols), along with the linear regressions of each sample as indicated in the legend; and mean $\pm1$ standard deviation bar plots for $m$ (b,e) and $b$ (c,f) for the complete and declustered catalogs, respectively. In the latter, the values in the parentheses corresponds to the coefficients of variation.}}
	\label{fig:random}
\end{figure*}

@@

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-09} 
	\caption{\myrevision{Variation of $<$$T_c$$>$, $k$-$M$ slope ($m$) and the $b$ value with respect to time for the seismic zone of Alborz for a moving window analysis of the visibility graphs of subsequences of $n=50$ consecutive events, along with the magnitude and the connectivity degree ($k$) of events in the 2016 period. Black triangles indicate the occurrence of significant earthquakes in each of the regions, with the corresponding magnitude at the top of each symbol.}}
	\label{fig:tc}
\end{figure*}

@@

\myrevision{In this windowing analysis, in addition to $b$ and $m$, we are interested in computing the value of $<$$T_c$$>$ explained in the Methodology section. Figure \ref{fig:tc} shows the earthquake sequence for the region of Alborz, and the variation of $k$, $m$, $b$, and $<$$T_c$$>$ with time. Although we performed the analysis for all three seismic zones, we concentrate in the region of Alborz because is the one with the larger number of events for the fixed-size moving window. We can see here how larger values of $k$ are associated with the larger magnitude earthquakes, and how the resulting values of $m$ for each 50-event sequence compares with $b$. In particular, we observe a good correlation between $m$ and $b$ fore the period prior to the 2010 $M_w$ 5.8 Damaghan earthquake, but after this event the correlation is not as clear. Unfortunately, contrary to the suggestions of \citet{Telesca2016}, we do not observe any clear correlation between the behavior of $<$$T_c$$>$ and the occurrence of the larger magnitude ($M_w \geq 5$) earthquakes in this region. Similarly, while the occurrence of larger events does seem to coincide with drops in $m$ and $b$, there are other similar changes in the values of $m$ and $b$ that are not associated with any significant event, making it difficult to draw strong conclusions. Nonetheless, in the region of Azerbaijan, we did observe more prominent changes in $m$ and $b$ with the occurrence of the 2012 $M_w$ 6.4 Tabriz earthquake, which leaves open the possibility of further study for regions with larger magnitude events.}

>>>>

We generate the visibility graphs for each subregion of interest using the time-magnitude sequences \myrevision{of both the complete and declustered catalogs. In the case of the complete catalog, that corresponds to the sequences shown in Fig.~\ref{fig:mag-time}. In every case, the numbers of nodes in each graph are the same as those shown in Table \ref{tab:seismicity}; and} the links between inter-visible events are established following the condition in equation (\ref{eq:vg}), as in the example shown in Fig.~\ref{fig:vg}. (We do not include a visualization of the graphs because the links are so many, that it is only practical to visualize the graphs of short sequences.) We collect information about the number of inter-visibility links associated with each event (connectivity degree, $k$) and categorize the events in magnitude bins of size $\Delta M_{\mathrm{bin}} = 0.1$, as mentioned in the Methodology section. Figure \ref{fig:km} shows the scattered distribution of events in the magnitude-connectivity degree plane for each subregion\myrevision{, corresponding to the complete catalogs.} The figure shows that, in general, the value of $k$ increases with $M_w$\myrevision{; and includes the results of obtaining linear $k$-$M$ regressions for each dataset along with the values of the slope $m$ for the three seismic zones, namely 10.05, 9.85, and 8.71 for the Azerbaijan, Alborz, and Kopeh Dagh regions. Similar results regressions were obtained for the declustered catalogs, which yielded $m$ values of 8.92, 8.97, and 7.96, respectively.}

@@

Next, we examine the relationship between the $b$ values from Table \ref{tab:seismicity} and the $k$-$M$ slopes. \myrevision{Figures \ref{fig:regression}a and \ref{fig:regression}b show the scattered results of the $m$ versus $b$ for the complete and declustered catalogs, respectively. Along with the data-points obtained for our analysis of the three seismic regions of northern Iran, in each plot, we also include the data points corresponding to the magnitude-time sequences of the Mexican subduction zone provided by \citet{Telesca2013}, the Pannoninan seismic zone from \citet{Telesca2014}, the experimental results obtained by \citet{Telesca2014-pone}, and the result from a similar preliminary analysis done by \citet{Azizzadeh_2017_SSA} for California. These figures also include the linear trends of the data points, as they are aggregated (i.e., each regression reflects the addition of new data-points from different studies). In the figure, we indicate which data points were considered for each of the computed linear regressions and their corresponding correlation coefficient, $R$.}

@@

To further explore the sensitivity of the graph properties to the number of events in each catalog and the value of the minimum magnitude, we randomly picked a significant number of subsequences from within the initial catalog compiled for each region, and repeated the analysis for each subsequence. In total, for each region, we extracted 200 new subsequences from the \myrevision{whole} catalog. The number of events in each subsequence, $n$, was varied randomly but chosen to be large enough to represent the seismic characteristics of each region. In particular, the minimum size of each subsequence was set to be \myrevision{$n \geq 200$}, and the maximum size in the sequence was set to be as large as the original \myrevision{whole} catalog (see Table \ref{tab:seismicity}). We forced the random subsequences to progress positively in time without altering the natural occurrence of events. In other words, we randomly determined the initial event and the subsequence size (number of events to be considered), and then picked that number of events following the initial earthquake in the subsequence. Next, we determined the value of $M_c$ and $b$ \myrevision{to obtain the complete and declustered catalogs for each region}, created the graph for all events with $M \geq M_c$, and extracted the connectivity degree of the events in each subsequence. \myrevision{Because this needed to be done for a large number of subsequences, in this part of the analysis we used the goodness-of-fit test (GFT) method to determine $M_c$ systematically, instead of the MAXC method. The GFT method follows the recommendations of \citet{Wiemer2000} and \citet{Wiemer2001}, and leads to less conservative values of $M_c$, which is preferable to keep the subsequences large enough.}

\begin{figure*}%[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-08} 
	\caption{\myrevision{Correlation between the $k$-$M$ slope ($m$) and $b$ values for the random subsequences extracted from the catalogs of the three northern Iranian seismic regions considered in this study (scattered symbols, 200 each region, 600 total) for both the complete (a) and declustered (d) catalogs, including the mean values (empty symbols with thick border) and the data points obtained from the regional analyses (solid symbols), along with the linear regressions of each sample as indicated in the legend; and mean $\pm1$ standard deviation bar plots for $m$ (b,e) and $b$ (c,f) for the complete and declustered catalogs, respectively. In the latter, the values in the parentheses corresponds to the coefficients of variation.}}
	\label{fig:random}
\end{figure*}

@@

\begin{figure*}%[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-09} 
	\caption{\myrevision{Variation of $<$$T_c$$>$, $k$-$M$ slope ($m$) and the $b$ value with respect to time for the seismic zone of Alborz for a moving window analysis of the visibility graphs of subsequences of $n=50$ consecutive events, along with the magnitude and the connectivity degree ($k$) of events in the 2005--2016 period. Black triangles indicate the occurrence of significant earthquakes in each of the regions, with the corresponding magnitude at the top of each symbol.}}
	\label{fig:tc}
\end{figure*}

@@

\myrevision{In this windowing analysis, in addition to $b$ and $m$, we are interested in computing the value of $<$$T_c$$>$ explained in the Methodology section. Figure \ref{fig:tc} shows the earthquake sequence for the region of Alborz, and the variation of $k$, $m$, $b$, and $<$$T_c$$>$ with time. Although we performed the analysis for all three seismic zones, we concentrate in the region of Alborz because is the one with the larger number of events for the fixed-size moving window. We can see here how larger values of $k$ are associated with the larger magnitude earthquakes, and how the resulting values of $m$ for each 50-event sequence compares with $b$. In particular, we observe a good correlation between $m$ and $b$ for the period prior to the 2010 $M_w$ 5.8 Damaghan earthquake, but after this event the correlation is not as clear. We do not observe any clear correlation between the behavior of $<$$T_c$$>$ and the occurrence of the larger magnitude ($M_w \geq 5$) earthquakes in this region. Furthermore, while the occurrence of larger events does seem to coincide with drops in $m$ and $b$, there are other similar changes in the values of $m$ and $b$ that are not associated with any significant event, making it difficult to draw strong conclusions. Nonetheless, in the region of Azerbaijan, we did observe more prominent changes in $m$ and $b$ with the occurrence of the 2012 $M_w$ 6.4 Tabriz earthquake, which leaves open the possibility to further study this for regions with larger magnitude events.}
